\OurTool{}: Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles\\[1em] Supplementary material

### 2019-12-22 Response-NAR-Round1.tex

	However, we have chosen to avoid an overly formal description of the method, as we believe such a writing style would not represent a good fit with the expectations of the typical NAR audience. We have adopted a linear narration of the method, which we completed in the hope to now ensure full reproducibility of the method.}

	\Comment{You do not introduce how you are able to compute MEA centroids without explicitely computing base pair probabilities. Checking your source code, I found the bp-probs are computed analogously to the structure probabilities, which is fine but should be explicitely mentioned to be reproducable.

		It is totally unclear to me (and I didnt took the time to track it down in your code), how you derive an MEA for a cluster with structures from different reactivity data sets, since the same structure can be part of the cluster multiple times (with different probe-constraint probabilites). Do you sum up probe-specific probabilities in that case? Do you derive new base-pair probabilities for the cluster (and if so how)?

	Please provide somewhere details of your methods concerning the whole bp-prob and MEA centroid computation (at least in the supplementary material). Otherwise, your method description leaves out a central point of your algorithmic pipeline and renders is non-reproducable!}

	\Answer{As mentioned in our response to the previous point, we revised our method description to include more details on the probability distribution on which we base our centroid/MEA structure. The computation of the MEA is performed using a strict reimplementation of the dynamic programming algorithm introduced by \cite{Lu2009}, so we just refer the reader to this article rather than reproducing its content.

	Please find below the description of the clustering method, as now appears in our revised document.
		\Quote{\paragraph{Clustering across conditions.} In order to infer recurrent conformations across sampled sets, \OurTool{} agglomerate structure (multi)sets while keeping track of their condition of origin, and \Def{clusters} with respect to the \Def{base-pair distance}, the number of base pairs differing between two structures. A clustering algorithm then partitions the (multi)set of sampled structures into \emph{clusters}, (multi)sets of structures such that the accumulated sum of distances over clusters is minimized.

				Among the many available options, we chose the Mini Batch k-Means algorithm~\citep{Sculley2010} (MBkM), implemented in the \Software{scikit-learn} pacakage~\cite{Pedregosa2012}, which requires less computational resources than the classic k-means algorithm, yet performed similarly in preliminary studies as an extensive collection of both agglomerative (affinity propagation) and hierarchical (Ward, Diana, McQuitty) clustering algorithms. A dissimilarity matrix, presenting the pairwise base-pair distance between structures, is precomputed and fed to the clustering algorithm.

				Any cluster $C$ output by the clustering is a multiset of structures, each labeled with its origin condition of $\mathcal{D}$.
				The \Def{cluster probability of a structure feature} $f$ (base pair or unpaired base) within a cluster $C$ is then defined as 
				$$ \mathbb{P}_C(f) = \frac{\sum_{\substack{S\in \mathcal{C}\\\text{s.t. } f\in S}} e^{-E(S)/RT}}{\sum_{\substack{S'\in \mathcal{C}}} e^{-E(S')/RT}} $$ 
				where $R$ represents the Boltzmann constant, $E(S)$ is the Turner free-energy, and $\mathcal{C}$ is the (non-redundant) set of structures in $C$. From those probabilities we define the \Def{centroid structure} of a cluster as its Maximum Expected Accuracy (MEA) structure, computed efficiently following Lu\etal~\citep{Lu2009}.

				Moreover, define the (pseudo-)\Def{Boltzmann condition probability} of a structure $S$, generated for a probing condition $d$ as part of a sampled  set $\mathcal{S}_d$, as
				$$\mathbb{P}_d(S) = \frac{e^{-E_d(S)/RT}}{\mathcal{Z}^*_d} \text{, with } \mathcal{Z}^*_d := \sum_{S'\in \mathcal{S}_d} e^{-E_d(S')/RT}$$
				where $E_d(S)$ is the pseudo free-energy assigned to the structure $S$ within the probing condition $d$.
				The \Def{stability} of a cluster $C$ denoted its accumulated pseudo-Boltzmann probability across conditions, computed as
				$$\text{Stability}(C) = \sum_{d\in \mathcal{D}} \sum_{S\in C\cap \mathcal{S}_d}  \mathbb{P}_d(S).$$
				A cluster is deemed \Def{significantly populated} if its stability exceeds a predefined threshold $\epsilon$. 
				We set  $\epsilon =\#{\sf Datasets}/3$ by default, such that at most three clusters are deemed significantly populated, and used as our primary candidates.
				Finally, we consider two clusters to be \Def{highly similar} if their centroid structures differ by at most $\delta$ base pairs ($\delta=1$ by default), allowing the identification of clusters in the presence of minor variations.

				The targeted number of clusters is a critical parameter of the \CL{} algorithm. It should, at the same time, remain small enough to ensure reproducibility, while being sufficiently large to discriminate outliers and ensure consistency within each cluster. We determine an \Def{optimal number of clusters} $\NumClust$ using an iterative heuristics, gradually increasing the number of clusters until a significantly-populated cluster is split into two similar clusters, or a poorly populated (outlier) cluster is created.
				Namely, our iterative heuristic consists in running \CL over an increasing number $\NumClust$ of clusters, starting with $\NumClust=2$, until the following \Def{stopping criterion} is met: 1) Two significantly populated clusters have associated centroids which are highly similar; or 2) Centroid structures of significantly populated clusters from the previous iteration are highly similar to those of the current iteration.}
}

	\Comment{Please replace "kT" by "\text{R}T" in the structure probability formula (and wherever needed)!}

	\Answer{We agree that using $k$ was unfortunate in this context, leading to an unnecessarily notational overload, and have used $R$ instead as suggested.}

@@

	\Comment{What is the difference betwee the "cummulative probability of [a cluster's] structures" (line 3-1-58) and the formally introduced "Stability(C)" notion? If there is none, please use accordingly, otherwise provide respective details.}

	\Answer{Those two notions indeed coincide and we had initially used a sentence to avoid a forward definition. We have now moved the stability definition earlier in the method description, and use it to define the notion of \Def{significantly populated cluster}.}

	\Comment{Please provide (eg in the supplement) some further details how to derived the values for m abd b for the Deigan method. So far, the description is quite vague. Within this context, you might want to discuss that too low pseudo-energy boni will yield similar results as hard constraints (which you want to avoid).}

	\Answer{The values $m$ and $b$ were loosely calibrated by grid search based on the \cite{Cordero2012} dataset, revealing a limited influence of those factors on the predictive performances. }

@@

	\Comment{The formula for "$Stability(C)$" misses a "$\sum_{S\in C \cap \mathcal{S}_d}$" between the sum over d and the probability.}

	\Answer{Absolutely, this is now fixed.}

	\Comment{Please put brackets around the probability sum within the "$Support(C)$" definition to make clear that you are comparing $\tau$ to the sum.}

	\Answer{Thanks for the suggestion, which greatly improves the readability of the formula.}

@@

	\Comment{If I get it right, if you are facing two clusters, one dominant in stability and the other in support, you will report two centroids, right?
	So reactivities that are inducing very high stability might dominate your prediction? Please check this hypothesis and if true discuss within the manuscript!}

	\Answer{
		The interpretation of the reviewer is accurate, and it is true that the configuration of reactivities may give rise to \Def{tight} (focused on a few structures), or \Def{wide} (compatible will multiple conformers) pseudo Boltzmann Ensembles. Combining a \emph{tight} and \emph{wide} pseudo-Boltzmann ensembles will grant more influence to the former over the final prediction. 
		In the worst case scenario, the combination of a poorly-informed tight Ensemble, which we informally call a \emph{bully}, with an accurate wide (\emph{open-minded}) Ensemble, may lead the bully to almost solely drive the ultimate prediction. However, we believe such cases to be rare, as indicated by the good performances of our approach on single probing experiments and that, in general, tight ensembles are indicative of successful experiments.

		Moreover, in the presence of multiple probing conditions, bullies can be overruled by a population of informed, even if open-minded, individuals. Our approach can indeed be seen, in first order approximation, as a weighted voting mechanism where each voter splits his vote between her/his preferred candidates. Since the probability of independently ill-informed bullies to agree by chance decreases exponentially with their number\footnote{In the absence of a systematic underlying bias, synchronizing the bullies, which in itself would be informative, or at the very least worthy of investigation.}, the voting process will ultimately favor a population of open-minded individuals over a, possibly majority, subpopulation of bullies.

		Finally, it is worth reminding that reactivity values are normalized, so that the average reactivity (over nucleotides touched by the probe) remains the same across conditions, so that the probability of a tight Ensemble being an artifact due to a given probing condition is expected to be limited. 
	}

@@

	\Comment{If you are NOT reporting multiple structures (as indicated in your results) how do you deal with cases illustrated above?}

	\Answer{
		Over a normal execution, \OurTool{} {\bf does report} all the dominant clusters (and even dominated ones), along with their centroids.
		In the context of benchmarking, unless explicitly specified, we systematically picked a cluster uniformly at random from the output to assess the predictive performances.

		However, in an overwhelming proportion of our experiments, a single cluster is reported, due to a generally good correlation  of stability with support. For instance, it is mathematically impossible for a single condition to feature more than one Pareto cluster, since the cluster $C^\star$ having max stability $s^\star$ is such that: either $C$ has support $t^\star=1$, in which case it dominates any other cluster with signature $(s<s^\star,t\in\{0,1\})$; or $C$ has support $t^\star=0$, ruling out the possibility of $t=1$ for any other cluster (since it would imply $s>s^\star$).

	\Comment{On page 7 you write "Fig. 5 suggests the presence of 8 clusters", which I do not agree... for me the PCA depiction would yield 4 clusters, i.e. B, C, D and the merge of all others. Thus, please rephrase the sentence or provide some other reason why you came up with 8 clusters from the beginning..}

	\Answer{We respectfully disagree, and still see 8 clusters in the PCA (and distance matrix, which we reformatted and reproduced below). Of course, this view is quite subjective and we carefully chose our wording (\emph{suggest}) to reflect the inherent limitations of a visual inspection. 
	\Quote{
			{\centering \includegraphics[width=.75\linewidth]{graphs/didy/bi_clustering}\\}%
		}
	On a more objective level, an examination of the distance matrix above indicates that merging conditions in the clusters A, E, F, G and H as suggested would be ill-advised, since some of the pairs of clusters ((E,F,G) vs H, A vs E\ldots) are actually quite distant. This  fact is somewhat obfuscated by the low-dimensionality of the PCA.
}

@@

	\Answer{We respectfully disagree, and still see 8 clusters in the PCA (and distance matrix, which we reformatted and reproduced below). Of course, this view is quite subjective and we carefully chose our wording (\emph{suggest}) to reflect the inherent limitations of a visual inspection. 
	\Quote{
			{\centering \includegraphics[width=.75\linewidth]{graphs/didy/bi_clustering}\\}%
		}
	On a more objective level, an examination of the distance matrix above indicates that merging conditions in the clusters A, E, F, G and H as suggested would be ill-advised, since some of the pairs of clusters ((E,F,G) vs H, A vs E\ldots) are actually quite distant. This  fact is somewhat obfuscated by the low-dimensionality of the PCA.
}

	\Comment{Concerning data normalization (Cordero et al.): when resetting all non-reacting position to 0, dont you loose the "background noise" of the experiment? 
	Did you test other options like subtracting the minimum/average/maximum of all non-reacting positions from all reactivities (before setting non-reacting to zero)?}

	\Answer{
		First, we need to stress that this cancellation of non-reactive nucleotides only applies to DMS and CMCT probing, and reflects the fact that those reagents are not supposed to react with certain nucleotides. Namely, DMS only reacts with \Ab{} and \Cb{}, while CMCT only reacts with \Gb{} and \Ub{}. We did verify that the reactivities associated with non-reacting nucleotides, as assessed using the relevant formula, are overwhelmingly close to zero so normalizing by the minimum/average/maximum would have no effect. Overall, these observations strongly support the idea of non-informative reactivity values at those positions. For instance, using them within the computation of the normalization (based on a percentile) would systematically increase the average reactivity, leading the pseudo-energy bonuses to act as \emph{hard constraints} for all practical purposes.

		Moreover, the reactivity normalization is based on percentiles over reacting positions, and there is no reason to suspect a systematic bias between the nature of nucleotides and the structural context. It follows that the effect of the normalization should be very similar when starting from a fully informed reactivity profile, or ignoring reactivities associated with \Gb{}/\Ub{} nucleotides.

		Finally, we must point out that we do not set the reactivities of non-reacting nucleotides to $0$, but rather to \emph{non-reactive}, usually encoded as a large negative value. This value is then interpreted as a missing data within \Software{RNAfold}, so that no pseudo-energy (rather than the bonus induced by a reactivity of $0$ by the \cite{Deigan2009} conversion) is used for those positions.

@@

}

	\Comment{I would be very interested to see the performance of multi-probe predictions via RNAfold (MFE+MEA) when merging their reactivity data.
	For instance: just average the reactivities (per position) and provide them as new pseudo-reactivity data to RNAfold.
	While this might extremely mess up predictions, it might also yield similar effects as IPANEMAP, i.e. importance of contradicting reactivities is reduced the more experiments are considered..}

	\Answer{
		In principle, this idea of averaging reactivities is natural, but has limits. 
		In particular, it does not sample from the same distribution as our approach, as can be seen in the following toy example:

		{\centering
			\begin{tabular}{ll}
				Structure $S^\star$ (free-energy $E^\star$)& \texttt{(((..)))(((..)))} \\
				Structure $S^\varnothing$ ($E^\varnothing=0$)& \texttt{................} \\
				Reactivity profile $r^\alpha$                          & \texttt{0011100000000000} \\
				Reactivity profile $r^\beta$                          & \texttt{0000000000011100}
			\end{tabular}\\}

		Now, consider a simplified ensemble of structures, consisting only of $S^\star$ and $S^\varnothing$, the open chain having free-energy $E^\varnothing = 0$. 
		Given a structure $S$ and a reactivity profile $r$, the overall pseudo-energy bonus is then 
		$$ \Delta G_r(S) = \sum_i m \log(r_i+1) + b.$$
		The energy bonuses for the combinations of structure/(averaged)reactivity profiles are

			$$
			\begin{array}{c|cc}
				\Delta G_r(S) & S:=S^\star & S:=S^\varnothing\\ \hline
				r:=r^\alpha & 2m\log 2 + 4\, b & 3m\log 2 + 16\, b\\
				r:=r^\beta & 2m\log 2 + 4\, b & 3m\log 2 + 16\, b\\
				r:=\frac{r^\alpha+r^\beta}{2} & 4m\log 3/2 + 6\, b & 6m\log 3/2 + 16\, b			
				\end{array}
			$$

		For instance, the emission probability of $S^\star$ given a reactivity profile $r$ is
		\begin{align*}\mathbb{P}(S^\star\mid r) &= \frac{\displaystyle e^{-\frac{\Delta G_{r}(S^\star) + E^{\star}}{RT}}}{\displaystyle e^{-\frac{\Delta G_{r}(S^\star) + E^{\star}}{RT}} + e^{-\frac{\Delta G_{r}(S^\varnothing)}{RT}} }\\
		&= \frac{1}{1 + e^{\frac{\Psi(R) + E^{\star}}{RT}}}
		\end{align*}
		with  $\Psi(r) := \Delta G_{r}(S^\star) -\Delta G_{r}(S^\varnothing)$.

		It follows that the expected frequency of occurrences of $S^\star$ within an overall multiset of $M$ structures sampled by \OurTool{} ($M/2$ structures for each condition) is given by 
		\begin{align*}
		\mathbb{E}\left(\frac{\#\text{Occ}(S^\star)}{M}\mid r^\alpha \wedge r^\beta\right) &= \frac{\mathbb{P}(S^\star\mid r^\alpha) + \mathbb{P}(S^\star\mid r^\beta)}{2}\\
		 &= \frac{1}{1 + e^{\frac{-m \log(2) -12\,b + E^{\star}}{RT}}} \\
		 &= \frac{1}{1 + \frac{e^{\frac{-12\,b + E^{\star}}{RT}}}{2^m}}
		\end{align*}
		while the same quantity, measured within $M$ structures sampled from an averaged reactivity profile $\frac{r^\alpha+r^\beta}{2}$, becomes
		\begin{align*}
			\mathbb{E}\left(\frac{\#\text{Occ}(S^\star)}{M}\mid \frac{r^\alpha+r^\beta}{2}\right) &= \frac{1}{1 + e^{\frac{\Psi\left(\frac{r^\alpha+r^\beta}{2}\right) + E^{\star}}{RT}}}\\
			  &=\frac{1}{1 + \frac{e^{\frac{-12\,b + E^{\star}}{RT}}}{(3/2)^{2m}}}.
		\end{align*}
		The two quantities are strictly different for any value of $m$ (except $m=0$), and the ratio of probability can go in different directions depending on the reactivities profiles and (exponentially large) ensembles of conformations.

		More empirically, we believe such a strategy would impact the robustness and versatility of our whole approach. 
		Indeed, in the presence of two sources of reactivities showing strong disagreement, indicating either the probing of two conformations, or an erroneous condition, averaged reactivity terms will end up favoring unpaired conformations, while our approach will probably produce two clusters (leaving aside the final Pareto optimization).
	}

@@

	\Comment{Given your tool's documentation, a couple of questions arise that should be answered within the README.

	\begin{itemize}\item How are the reported free energies evaluated since your predictions are based on (multiple) reactivity-biased predictions? (your code hints at RNAeval calls, but please be verbose in the documentation to enable a clear intepretation of the results and comparability with other tools' output)\end{itemize}}

	\Answer{Free-energies are re-computed using the \Software{RNAeval} tool, we have completed the documentation to make this point more explicit (and refer to a refined definition of these terms in the manuscript).}

@@

	\Comment{\begin{itemize}\item How are the reported Boltzmann probabilities computed? based on the overall sample set? the respective cluster? (in both cases: what energies are used?) or based on the true (unconstraint) partition function computed e.g. by RNAfold?\end{itemize}}

	\Answer{The reported Boltzmann probabilities, also called Stability in the manuscript, are strictly computed as described in the manuscript. We have made this point explicit in our revised documentation.}

@@

	\Comment{Given Table 1 (CMCT) and the general unawareness of the Nussinov-like MEA-optimization of the importance of base pair stacking, you might be able to further improve your results when implementing an MEA variant that excludes lonely base pairs.}

### 2019-12-22 Response-NAR-Round1.tex

	\Answer{We agree that this would be a potentially fruitful development. This would require a modification of the dynamic programming equations presented by \cite{Lu2009}, for instance by leveraging the Waterman grammar into something similar to:
	\begin{align*}
	  S &\to \texttt{(}\texttt{(} \,T\, \texttt{)}\texttt{)}\, S \mid \texttt{.}\, S \mid \varepsilon \\
	  T &\to \texttt{(}\, T\, \texttt{)} \mid \texttt{(}\texttt{(} \,T\, \texttt{)}\texttt{)}\, U\mid   \texttt{.}\, S \\
	  U &\to \texttt{(}\texttt{(} \,T\, \texttt{)}\texttt{)}\, S \mid \texttt{.}\, S 
	\end{align*}
	Such a modification would triple the memory requirement, but would remain polynomial in time and should not constitute the bottleneck of the whole method. However, it would also warrant an extensive validation effort, and lead to the description of further material within an, already consequent, Methods section. For these reasons, we prefer to leave this promising extension to a future version of \OurTool.}

@@

	\Comment{I highly recommend to integrate IPANEMAP into the package management system conda via the bioconda channel. This would:
		\begin{itemize}
	\item be easy since IPANEMAP is eventually a set of python scripts (only a simple build and dependency script needs to be integrated into bioconda's recipes github repo)
	\item  handle all dependencies for the user
	\item ensure a simple installation of the tool and all dependencies 
	\item increase visibility of the tool
			\end{itemize}}

	\Answer{Again, we agree that integrating the package within \Software{conda} would be ideal. However, since the \Software{ViennaRNA} package is not currently available on \Software{bioconda} for all platforms (only Mac OS and Linux), this would still require a manual installation and would somewhat mitigate the benefits of such an integration.

		In the new version of the documentation, we at least partially ease the installation process, and suggest using \Software{pip} (available in any \Software{Python} release) to install all the necessary dependencies through the command:

		{\centering \Software{pip install cython scipy numpy sklearn}\\[.5em]}

		The \Software{ViennaRNA} package, however,  still needs to be independently installed. We have contacted the maintainers of the \Software{ViennaRNA} package to (help) make the package more universaly accessible, and will certainly pursue the  integration of future versions of \OurTool{} within \Software{bioconda} (a port to \Software{Python3} is currently underway).
	}

@@

	\Comment{When reading your work, it reminds me of another application, ie. the computation of unpaired probabilities for RNA-RNA interaction prediction in the presence of multiple probing data. For instance, see 
	https://doi.org/10.1093/bioinformatics/bty1029
	where single-experiment-reactivities were integrated (via Vienna package routines). 
	Given your pipeline, you should be able to derive suitable probabilities either directly from Pareto-optimal cluster(s) or from the overall sample set.}

	\Answer{This is a great suggestion for a future application of our method, and will rank high within the list of our priorities (along with the, arguably natural analysis of multi-states RNAs). However, we feel that such a study would warrant a validation effort that would go beyond the scope of the current paper.}

	\Comment{Fix the following typos/unfortunate choices of words:\begin{itemize}
	\item general: et al $\to$ et al. \IAnswer{Done}
	\item 1-2-30 : please provide the abbreviation details for SHAPE \IAnswer{Done}
	\item 1-1-21 : remove comma after "experiments" \IAnswer{Done}
	\item 1-2-37 : "more dynamics"\IAnswer{Done}
	\item 1-2-49 : "qualitative"\IAnswer{Done}
	\item 1-2-53 : "of individual nucleotides"\IAnswer{Done}
	\item 2-1-23 : "These include " beside structure optimization " the joint folding.."\IAnswer{Done}
	\item 2-1-49 : signal $\to$ signals\IAnswer{Done}
	\item 2-2-38 : criterion "are" $\to$ "is"\IAnswer{Done}
	\item 2-2-39 : identified $\to$ represented\IAnswer{Done}
	\item 3-1-5 and 8 : better replace "$R_i$" with "$d_i$", since you are using R for reference later on \PAnswer{We replaced $R_i$ with $r_i$ to avoid clashing with the new notation for the Boltzmann constant $R$. We also denote the reference structure as $\Ref$ in the definitions of the evaluation metrics.}
	\item 3-1-60 : maybe replace "\#Datasets" with "$|\mathcal{D}|$" ? \IAnswer{Done}
	\item 3-2-17 : missing "are" in "that both" \IAnswer{Done}
	\item 3-2-32 : move "where $k$ is the number of clusters" to line 40, where it is needed. \IAnswer{Done}
	\item 4-1-8 : Within the formula "$Dist(d,d')$" you have to replace "$x$" with "$i$" \IAnswer{Done}
	\item 4-1-52 : "Non-canonical" $\to$ "Crossing" or "Pseudoknot-forming" \PAnswer{We clarified this description, which originally mixed the notions of non-canonical BPs and pseudoknots in a confusing manner. The new sentence reads \Quote{Non-canonical base pairs were removed, and a non-pseudoknotted secondary structure was extracted as a maximum subset of non-crossing base pairs~\citep{Smit2008}.}}
	\item 5-1-40 : "metrics" $\to$ "metric" \IAnswer{Done}
	\item 6-2-44 : "hajdin" $\to$ "Hajdin" \IAnswer{Done}
	\item 8-1-31 : "Figure 1" $\to$ "6" \IAnswer{Done}
	\item 10-1-56 : "Figure" $\to$ "Table" \IAnswer{Done}
		\end{itemize}}	
	\end{enumerate}

	\newpage
	\subsection{Reviewer \#3}

	\begin{enumerate}[resume]
	\Comment{My primary concern is that the authors use a pseudo-free energy formulation (page 3) with parameters that are the same for all conditions and probing reagents.  I am very surprised that this works.  The \cite{Deigan2009} work fit the parameters empirically for SHAPE data.  \cite{Eddy2014} more explicitly drew the connection between these parameters and the log odds ratio of a nucleotide being paired, following earlier work of \cite{Suekoesd2013}.  \cite{Cordero2012} also explicitly fit the log odds for DMS data using gamma distributions.  I would expect that each probing reagent and each probing condition would need different parameters.\label{arg:homogenous}}

		\Answer{We also expect that additional conversion formulae, and a specific parameter calibration, for each condition, would be beneficial to the downstream predictive performances. In fact, it is one of our ongoing (recently funded) projects  to produce more extensive data for a subset of the conditions explored in this study, and to attempt to systematically learn reactivities-to-pseudo-energy conversion models.

		However, the conjunction of:
		\begin{itemize}
			\item Distinguishable, yet large-variance and highly-overlapping, reactivity distributions being observed for unpaired and paired nucleotides;
			\item High reproducibility for  reactivity profiles across repeats (and, albeit less frequently, across probes as shown in Figure A above);
		\end{itemize}
		supports the notion that the determinants of SHAPE reactivity extends far the current binary paired/unpaired model. 
		This hypothesis is already explored to some extent in \cite{Suekoesd2013}, and confirmed by recent mechanistic studies~\citep{McGinnis2012,Sexton2017,Hurst2018,Mlynsky2018,Frezza2019,Busan2019}.

		This leads us to a strong suspicion that future pseudo-energy models will have to consider more structural contexts than the current binary one. Those contexts will need to be captured by the dynamic programming schemes. The training of such extended models would probably induce a highly heterogeneous quality for pseudo-energies models trained across conditions, given  the relative scarcity of available data for certain reagents/technologies. In particular, training extended parameters would probably lead to overfitting in contexts where only a couple of reactivity profiles are currently available.

		Moreover, going back to the suggested use of the gamma distribution in the case of DMS, it is worth pointing out that \cite{Cordero2012} state in their supplementary material that
		\Quote{
		The performance of the algorithm using this probabilistic potential for SHAPE, DMS, and CMCT reactivities is given in Table S6 and is identical to
		results obtained for the standard potential with slope ( m ) and intercept b ) optimi zed through grid
		search .}
		We also note the relative stagnation of performances of state-of-the-art tools integrating SHAPE data, despite an increasing sophistication of the pseudo-free-energy models. So, while we believe progress can be made in the integration of reactivities, we do not expect them to radically change the quality of predicted models, nor to invalidate the findings of this manuscript.

		To summarize, it seems too early in this manuscript to pursue a specialization of energy models to each reagent/probing technology, and the benefits of considering comparable reactivity profiles (produced in an homogeneous manner, and similarly normalized) far outweigh the expected loss in predicting accuracy. Consequently, rather than specialize our treatment of individual probes and address this heterogeneity as a possible confounding factor in our analyses (in a way that would remain to be determined), we prefer to preserve a homogeneous treatment for all sources of probing data.

@@

		}

	\Comment{The manuscript needs to be clearer about how the two parameters (m and b) determined for this work.  What optimizations were performed?  It needs to be clear that proper cross-validations were performed if the parameters were optimized.  Conversely, if the optimizations were not performed, it seems like an opportunity was then lost to have the best accuracy.}

		\Answer{
			The (hyper)parameters $m$ and $b$ were initially calibrated on the \cite{Cordero2012} dataset, revealing a very limited impact of the parameters on the prediction accuracy. We kept those values for multi probing, as the scarcity of available data did not allow us to afford an extensive cross-validation. We updated the description as shown below to be more explicit.
			\Quote{
			Those values were halved in comparison to those recommended by Deigan\etal\citep{Deigan2009}, following both a grid search optimization on the Cordero dataset, and based on the rationale that lower absolute values for pseudo-energy bonuses increase the expected overlap between pseudo-Boltzmann ensembles.}
		}

@@

	\Comment{I am also concerned that the manuscript is not as clear as it should be about the three types of multiple data that are used.  It can use multiple reagents, probing data for multiple solution conditions, or probing data across multiple sequence variants (a.k.a. mutate and map data).  It is a real achievement that one program handles all three cases.  The abstract does state this (briefly).  The Introduction is not clear on this points, just calling the data “multiple probing data.”  Also, the text itself is sometimes unclear about what is meant.  On page 6, for example, the sentence “It is also worth mentioning that NMIA+CMCT, combining the best and worst conditions, achieves a better combined performance than DMS+NMIA, the two best mono-probing conditions.” calls these “multiple conditions”, but these data are from multiple probing reagents.  I would like to see a revised manuscript be clearer about the three types of information in the Introduction.  In the text, multiple reagents, multiple solution conditions, and multiple sequence variants should be clearly denoted as different multiple data types. }

	\Answer{\TODO{Clarify in manuscript what conditions means in the context of the study}}

	\Comment{My third concern is that IPANEMAP is not entirely stable, in that there are variations in the results from run to run.  Users of the software will want the overall results to be stable; it is hard for most users to use a program multiple times and integrate the results.  It seems to me that the stability would be improved by using larger sample sizes, and that this should be explored for this manuscript.  If larger sample sizes don’t improve the stability, this should be addressed. }

	\Answer{
		We agree that stability of the predictions is potentially problematic, especially in the context of reproducing earlier results.

		Larger sample sizes represent a natural option for stabilizing the predictions, and indeed seem to decrease the variability. However, a substantial increase of the number $x$ of samples induces extreme computational demands, due to the current necessity to compute the dissimilarity matrix in $\Theta(x^2.n)$ time and $\Theta(x^2)$ memory. In this context, simply multiplying by 10 the number of samples would induce a 100 fold increase of the time requirement. In the future, we will revisit the clustering component to learn suitable embeddings for secondary structures, granting access to linear-time EM algorithms (eg based on Gaussian Mixture Models).

		\Quote{
	{\centering \includegraphics[width=.7\linewidth]{graphs/boxplotreproducibility}\\}
}					
		Moreover, as noted by this reviewer, an expert user would ensure a greater stability by running \OurTool several times (say 10), and electing the most recurrent structure. This process would achieve asymptotic determinism, since the frequency of a structure would converge towards real emission probabilities. Moreover, most frequent structures seem to be the best quality ones, and are emitted by \OurTool with fairly large probability, ensuring convergence in a reasonable number of iterations.
		However, such a process would be cumbersome for a basic user.

		In the future, for the sake of reproducibility, we also wish to report, and letting the user specify, a \Def{random seed}, which will  allow the exact reproduction of an analysis. However, such a feature is currently not supported by \Software{RNAsubopt}, so we are working towards its inclusion to the \Software{ViennaRNA} package.

@@

	}

	\Comment{On page 3, the text states that “Non-canonical base pairs were removed (40), by considering as the secondary structure the maximum subset of non-pseudoknotted base pairs.” I think the manuscript should state that non-canonical pairs were removed and that pseudoknots were also removed.  Just removing the pseudoknots would not remove the non-canonical pairs. }

	\Answer{This is entirely correct. We clarified this description, which originally mixed the notions of non-canonical BPs and pseudoknots in a confusing manner. The sentence now reads: \Quote{Non-canonical base pairs were removed, and a non-pseudoknotted secondary structure was extracted as a maximum subset of non-crossing base pairs~\citep{Smit2008}.}}

@@

	\Comment{On page 5, the MCC should be stated as the compromise between sensitivity and PPV (not sensitivity and specificity).}

	\Answer{While the statement was meant to be informal, we agree (and it is quite established in the field) that the sensitivity is not a relevant metric for assessing the predictive capacities of structure prediction methods, and that the MCC correlates very well with the product of the Sensitivity and PPV. We have decided to preserve the informal tone, and now state: \Quote{It can be interpreted as a compromise between the main metrics derived from the confusion matrix, and is defined as [..] }}

	\Comment{Figure 2 shows the accuracy of structure prediction using NMIA, DMS, and/or CMCT data.  As noted above, the pseudo free energy parameters for SHAPE (NMIA) will not be optimal for DMS and CMCT.  The benchmarks should therefore also be run using the pseudo free energies fit by the Das lab for DMS and CMCT~\citep{Cordero2012}.  My hypothesis is that the structure prediction accuracy will be higher with the properly fit parameters.}

	\Answer{
		Unfortunately, due to our current reliance on \Software{RNAfold}, which takes reactivities as input rather than direct pseudo-energy terms, we could not test this hypothesis. However, we do not expect such a fitting procedure to produce substantially better results. Indeed, as stated in the supplementary material of the \cite{Cordero2012} manuscript:
		\Quote{The performance of the algorithm using this probabilistic potential for SHAPE, DMS, and CMCT reactivities is given in Table S6 and is identical to results obtained for the standard potential with slope (m) and intercept (b) optimized through grid-search.}
		While this leaves open the possibility of improvements upon re-optimization of the $m$ and $b$ parameters, this statement indicates that the nature of the conversion itself does not play a large role in the final performances. 
	}

	\Comment{In the Conclusion and Discussion, it is an oversimplification to state that Rsample also uses stochastic sampling.  Rsample attempts to adaptively define the pseudo free energies to not perturb probabilities for nucleotides in an initial stochastic sample that are consistent with the SHAPE mapping data.  Perhaps this manuscript should also mention that detail and state this adaptive approach is possibly not as accurate (when producing a single structure) as the \cite{Deigan2009} approach.}

	\Answer{
	This refined free-energies could indeed be impacting the predictive capabilities of \Software{RSample} (although their underlying rationale is convincing), we came to realize upon further inspection that \Software{RSample} actually never uses sampling/clustering on the \cite{Hajdin2013} dataset. Rather, the method concludes from the corrective factors that a single conformation is most likely, and simply returns the MEA of the whole pseudo-ensemble. We have updated our discussion to include this alternative explanation, which seems to explain the strong discrepancy, and supports the notion that clustering can be used to \emph{clean up} a noisy Boltzmann ensemble.
	\Quote{
In this context, the good performance of \OurTool in comparison with \Software{RSample}~\citep{Spasic2017} may appear surprising. Indeed, while independently developed and used~\citep{Deforges2017}, both methods share a common sampling/clustering scheme. However, \Software{RSample}~\citep{Spasic2017} reverts to a MEA~\citep{Lu2009} computation whenever a single conformation is suspected. The MEA structure coincides with the centroid of the whole Boltzmann ensemble, suggesting that the increased performances of \OurTool could be at least partially attributed to the use of sampling, and to our adaptive choice of the number of clusters.
Moreover, even in cases where a single dominant conformation is expected, considering multiple clusters seems to create more opportunities for discarding noisy random unstable conformations. Such conformations naturally occur in a Boltzmann-distributed subset of structures, and may end up \emph{polluting} the centroid structures by artificially inflating the cluster diversity. }}

	\Comment{Typos, etc.:
		\begin{itemize}
		\item The quality of writing is excellent.  There are few typos and the text is clear.  
		\item I noted in the second column on page 3 that k is defined before the equation that uses it (1/(1+k)).  It would be more conventional to define k after the equation.  (I thought I had missed something as I read through).  \IAnswer{Done}
		\item In the Funding section, “phD” should be “PhD”. \IAnswer{Done}
		\item There are many problems with the references, however.  Many references are incomplete (missing page and volume, e.g. 22, 23, …).  A number of references also have superfluous months with the years (e.g. 5, 6, 7,…). \PAnswer{We have extensively sanitized our references, and we believe our bibliography now adheres to accepted standards.}
		\item It would be helpful if the caption to figure 2 named the benchmark dataset.  This is clear in the main text, but will make it easier to quickly digest the manuscript if the captions are all self-contained.\PAnswer{We have systematically revised the captions to make the as self-contained as possible, and in particular always name the dataset on which the data was produced.}
	\end{itemize}
}

	\end{enumerate}
\newpage

@@

	\subsection{Reviewer \#4}

	\begin{enumerate}[resume]

	\Comment{IPANEMAP is benchmarked against Rsample in a mono-probe setting. It is shown to outperform Rsample, where the authors attribute this to the sample denoising IPANEMAP performs via clustering (page 7). I am not sure if this is indeed the case and also think this benchmark could be improved in several ways.}

	\Answer{First of all, we would like to point out that the performances of \OurTool{} in a mono-probing setting were explored as a validation of the basic predictive capacities of \OurTool{} in a context where reference performances are available. The very good behavior of \OurTool{} on this benchmark came as a pleasant surprise which we deemed worthy of a mention in the paper.

	In particular, this benchmark should not be mistaken for a central result in our paper, and we do not claim that \OurTool{} supersedes \Software{RSample}. Indeed, the benchmark on single-state RNAs is not the most salient point of the \cite{Spasic2017} paper, since the main application of \Software{RSample} is to identify multiple conformations, a use-case that we did not specifically tried to address while designing \OurTool{}.}
